Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-29 review_responses.tex

\subsection{R1.7 Provide Practical Takeaways}
\label{R1.7}
\noindent\textbf{Review:} \\
\noindent \textbf{Provide practical takeaways or usage guidelines for different loss-synth combinations. } 
\\

\noindent\textbf{Response} \\

In \textbf{IV.A Key Findings}, we discuss important findings and provide answers to our research questions. The following section, \textbf{IV.B Practical Recommendations}, distills our findings into actionable heuristics for practitioners.

\begin{displayquote}
    Our findings not only provide answers to the research questions proposed in the introduction, but also serve as guidelines for practitioners in selecting appropriate similarity measures and synthesis methods for future sound-matching experiments.
\begin{itemize}
    \item In lieu of listening tests, MSS or P-Loss are generally useful for large-scale benchmarking. However, we encourage confirming the results with manual listening, particularly if specificity is important. 
    \item For amplitude-modulated synthesis, DTW-based losses such as \DTWEnv{} are recommended over spectrogram differences and JTFS.
    \item  For subtractive/noise-filtered synthesis, spectrogram based losses are most effective. Based on our results, SIMSE appears to be the better method of spectrogram comparisons relative to L1, but the amplitude agnosticism of SIMSE likely plays a role in its poor performance in additive synthesis.
    \item Iterative differentiable optimization via the Faust-to-JAX pipeline is a viable strategy for defining various differentiable DSP functions, requiring only modest hardware resource and providing a more natural approach to synthesizer definition.
    \item In our amplitude modulation experiments, \JTFS{} did not outperform other measures, contradicting prior claims of its superiority for mesostructures~\cite{vahidi2023mesostructures}. Thus, our results call into question its effectiveness as a general mesostructure measure, and highlight a need for further research into its utility.
\end{itemize}

@@

\subsection*{R2.3 Clear Takeaways}
\noindent\textbf{Reviewer:}\\
\textbf{This work addresses the problems on ``Loss Selection'' and ``Synthesis Selection.''  However, the results as presented here are largely inconclusive (i.e. it is not imminently clear how to select losses/synthesizer pairings), leaving very little in the way of clear takeaways sparking future research.\\}

\noindent\textbf{Response:} \\
We acknowledge the reviewer’s concern that our main result may appear inconclusive. However, we see this as an important finding in itself: much of the prior literature implicitly over-claims general effectiveness of particular loss functions based on narrow or in-domain settings. Our revisions now explicitly emphasize that the lack of a “silver bullet” for loss functions is a core contribution of this work. By showing that loss effectiveness depends on the synthesis context, the inconclusiveness can be seen as evidence against over-generalization, and an argument for more thorough, context-sensitive benchmarks in measuring sound similarity.\\

As for clear takeaways, we've added two new sections which address the valid issues raised here and by Reviewer 1.

\begin{itemize}
  \item \textbf{IV.B Practical Recommendations} which distills our findings into actionable guidelines for practitioners. DTW\_Env is recommended for amplitude-modulated synthesis, SIMSE for subtractive/noise-filtered synthesis, and MSS or P-Loss as large-scale proxies for sound-matching evaluation (with manual validation being strongly advised). We also note that our results challenge prior claims about JTFS’s universal superiority for mesostructures.
  \item \textbf{IV.D Applicability to More Complex Synthesizers} that explicitly discusses how the insights from our controlled two-parameter DSP-based synthesizers generalize to richer synthesizers with many parameters and complex routing. While acknowledging the limitations of our simplified testbed, we argue that the observed dependencies between synthesis method and loss function remain informative even in more complex settings.
\end{itemize}
For a more detailed response, see our response to the Reviewer.1 in Sections~\nameref{R1.6} and \nameref{R1.7}. Our response to \nameref{R1.5} may also be of interest.

### 2025-09-29 sections/background_related.tex
Updates from Overleaf

A neural approximation for another part of the supervised sound-matching chain---this time the synthesizer---was proposed by Barkan \textit{et al.}~\cite{barkan2023inversynthII}. As they noted, without a differentiable synthesizer, ``model-based'' (or what we call supervised) approaches cannot directly compare $\xtheta$ to $\xzero$, often opting for P-Loss, which may not correctly map the parameters of sound to the output audio~\cite{esling2019flow,han2023perceptual,masuda2023improving}. This approach used a model mapping sounds to parameters, another approximating the synthesizer, and a loss function which combines P-Loss with STFT differences. This ``Inversynth II'' (IS2) approach yielded significant improvements to previous works which did not use the STFT approximations for loss~\cite{esling2019flow,barkan2019inversynth}. Barkan \textit{et al.} then attempted to improve the IS2 model with Inference-Time
Fine-tuning (ITF). For ITF, the synthesizer approximation is frozen and the encoder is iteratively fine-tuned for a particular sample. However, ITF often degraded performance, possibly due to proxy–synthesizer mismatch or overfitting.

@@

 Audio embeddings can also be used as a similarity metric. In a recent work, Cherep \textit{et al.} used latent representations from the CLAP model~\cite{wu2023large} along with a differentiable synthesizer~\cite{synthhaxcherep2023} to create creative interpretations of sound effects. In their approach, a desired sound-effect is described in text and embedded using CLAP, followed by iterative updates from a gradient-free optimizer~\cite{evosax2022github} to the synthesizer's parameters to minimize embedding differences. Based on manual hearing tests, this approach did not produce the ``correct'' sounds more frequently than previous works~\cite{kreuk2022audiogen}. However, it did yield better scores for ``artistic-interpretation'' (or what this works refers to as ``imitation'').

@@

Having looked at the current literature, we identify four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two issues are left for future work.
\begin{enumerate}
    \item \LossSelect: It remains unclear whether there exists a universally best-performing loss function, 
    or whether performance depends on factors such as the sound domain, synthesizer architecture, 
    and desired output characteristics. Existing studies typically evaluate losses only with simple synthesis 
    setups~\cite{vahidi2023mesostructures}, leaving open questions about their generality. 
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The effects of using different isolated DSP functions in iterative sound-matching and how these interact with loss functions remains largely untested.
    \item \PeriodicLoss: Loss function landscapes are not easy to navigate, as there can be many local minima, or large flat areas~\cite{turian2020sorry,vahidi2023mesostructures}.  In differentiable settings, this can cause gradient descent updates to not reach the global minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is an under-explored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

@@

Together, these design choices supply the controlled experimental evidence missing from the literature, 
clarifying how losses and synthesis methods interact in differentiable, iterative settings, 
and offering insights likely to generalize to more complex domains.

>>>>

A neural approximation for another part of the supervised sound-matching chain---this time the synthesizer---was proposed by Barkan \textit{et al.}~\cite{barkan2023inversynthII}. As they noted, without a differentiable synthesizer, ``model-based'' (or what we call supervised) approaches cannot directly compare $\xtheta$ to $\xzero$, often opting for P-Loss, which may not correctly map the parameters of sound to the output audio~\cite{esling2019flow,han2023perceptual,masuda2023improving}. This approach used a model for mapping sounds to parameters, another for approximating the synthesizer, and a loss function which combines P-Loss with STFT differences. This ``Inversynth II'' (IS2) approach yielded significant improvements to previous works which did not use the STFT approximations for loss~\cite{esling2019flow,barkan2019inversynth}. Barkan \textit{et al.} then attempted to improve the IS2 model with Inference-Time
Fine-tuning (ITF). For ITF, the synthesizer approximation is frozen and the encoder is iteratively fine-tuned for a particular sample. However, ITF often degraded performance, possibly due to proxy–synthesizer mismatch or overfitting.

@@

 Audio embeddings can also be used as a similarity metric. In a recent work, Cherep \textit{et al.} used latent representations from the CLAP model~\cite{wu2023large} along with a differentiable synthesizer~\cite{synthhaxcherep2023} to create creative interpretations of sound effects. In their approach, a desired sound-effect is described in text and embedded using CLAP, followed by iterative updates from a gradient-free optimizer~\cite{evosax2022github} to the synthesizer's parameters to minimize embedding differences. Based on manual hearing tests, this approach did not produce the ``correct'' sounds more frequently than previous works~\cite{kreuk2022audiogen}. However, it did yield better scores for ``artistic-interpretation'' (what we referred to as ``imitation''). It should be noted that audio embedding models are often trained using simpler loss functions---particularly those which utilize Fourier transforms---therefore improving such simpler loss functions is likely to improve the embedding models.

@@

Having looked at the current literature, we identify four major areas of weakness in sound-matching. We target the first two issues in this work, while the latter two are left for future work.
\begin{enumerate}
    \item \LossSelect: It remains unclear whether there exists a universally best-performing loss function, 
    or whether performance depends on factors such as the sound domain, synthesizer architecture, 
    and desired output characteristics. Existing studies typically evaluate losses only with simple synthesis 
    setups~\cite{vahidi2023mesostructures}, leaving open questions about their generality. 
    \item \SynthSelect: There is a lack of diversity in the synthesis methods used in sound-matching. The effects of using different isolated DSP functions in iterative sound-matching and how these interact with loss functions remains largely untested.
    \item \PeriodicLoss: Loss function landscapes are not easy to navigate, as there can be many local minima, or large flat areas~\cite{turian2020sorry,vahidi2023mesostructures}.  In differentiable settings, this can cause gradient descent updates to not reach the global minima. 
    \item \OutDomain: Sound-matching with out-of-domain sounds is an under-explored area, yet a necessary one for practical applications for sound designers.
\end{enumerate}

@@

Together, these design choices supply the controlled experimental evidence missing from the literature, 
clarifying how losses and synthesis methods interact in iterative sound-matching,
and offering insights likely to generalize to more complex domains.

### 2025-09-29 sections/methodology_and_analysis.tex
Updates from Overleaf

Here we discuss the implementation of the four loss functions (Section~\ref{sec:loss_implementation})
across four differentiable synthesizers (Section~\ref{sec:programs}). For
each loss–synthesizer pair, we run a large number of trials and evaluate
their outcomes automatically and manually
(Section~\ref{sec:evaluation_manual_auto}). ``Experiments'' in this context
refer to one complete iterative sound-matching run, from random parameter
initialization through gradient-based optimization until termination. Scores
from these experiments are then aggregated and statistically compared to
determine best-performing losses and synth–loss pairings.

>>>>

Here we discuss the implementation of the four loss functions (Section~\ref{sec:loss_implementation})
across four differentiable synthesizers (Section~\ref{sec:programs}). For
each loss–synthesizer pair, we run a large number of trials and evaluate
their outcomes automatically and manually
(Section~\ref{sec:evaluation_manual_auto}). ``Experiment'' in this context
refers to one complete iterative sound-matching run, from random parameter
initialization through iterative gradient-based optimization until termination. Scores
from these experiments are then aggregated and compared to
determine best-performing losses and synth–loss pairings.

### 2025-09-29 sections/summary_weakness_conclusion.tex
Updates from Overleaf

\textbf{Q1}: To what extent do automatic evaluation metrics agree with manual listening tests? 
\\We see somewhat consistent results between the rankings assigned by manual hearing tests (which we take as the ground truth), and automatic measures of P-Loss and MSS. With the exception of P-Loss in \BPNoise{} narrowly selecting a different spectrogram based loss, all measures of performance selected the same top performer. However, given the simplicity of our programs and the occasional deviations, we cannot conclude that these automatic measures are substitutes for human hearing tests. As many previous works have done (see Table~\ref{tab:summary}, the use of MSS and/or P-Loss as general loss functions does seem appropriate, however, we advise the use of listening tests or custom made loss functions for more specific analysis.

\textbf{Q2}: Are DTW and SIMSE effective measures of loss? If so, in what contexts?\\
\DTWEnv{} consistently outperformed other measures in amplitude-modulated synthesis, where envelope alignment is critical. \SIMSESpec{} performed best in subtractive synthesis, where scale-invariance captures noise-filtering effects more robustly than L1-based measures. These results suggest that the advantages of these novel losses are context-dependent, underscoring the need for further exploration of creative differentiable similarity measures.

\textbf{Q3}: Can iterative differentiable optimization be applied effectively in synthesizers using classical DSP functions? \\
The approach of designing DSP functions in Faust and transpiling to differentiable Faust code yielded meaningful outputs and enabled comparative evaluation of losses. The use of Faust for synthesizer definition was convenient, and transpiling to Jax enabled quick gradient optimization. We were able to run our 1200 experiments within 72 hours on a laptop without a GPU\footnote{Lenovo ThinkPad T480, i5-8350U CPU at 1.70GHz, 32 GB RAM.}.

\subsection{Practical Recommendations}
    Our findings not only provide answers to the research questions proposed in the introduction, but also serve as guidelines for practitioners in selecting appropriate similarity measures and synthesis methods for future sound-matching experiments.
\begin{itemize}
    \item In lieu of listening tests, MSS or P-Loss are generally useful for large-scale benchmarking. However, we encourage confirmation of conclusions with manual listening, particularly if specificity is important. 
    \item For amplitude-modulated synthesis, DTW-based losses such as \DTWEnv{} are recommended over spectrogram differences and JTFS.
    \item  For subtractive/noise-filtered synthesis, spectrogram based losses are most effective. Based on our results, SIMSE appears to be the better method of spectrogram comparisons relative to L1, but the amplitude agnosticism of SIMSE likely plays a role in its poor performance in additive synthesis.  
    \item Iterative differentiable optimization via the Faust-to-JAX pipeline is a viable strategy for defining various differentiable DSP functions, requiring only modest hardware resources and providing a more natural approach to synthesizer definition.
    \item In our amplitude modulation experiments, \JTFS{} did not outperform other measures, contradicting prior claims of its superiority for mesostructures~\cite{vahidi2023mesostructures}. Thus, our results call into question its effectiveness as a general mesostructure measure, and highlight a need for further research into its utility.
\end{itemize}

@@

The main hypothesis tested here is whether the performance of differentiable loss functions is influenced by the synthesis techniques used for sound-matching. We conducted systematic iterative sound-matching experiments by combining four different loss functions with four different sound synthesis programs. We ranked the performance of the iterative sound-matching pipelines for every loss function and program, and observed that the success of the pipeline (that is, how closely the output sound matches the target sound) is program dependent. In other words, different synthesizer programs work best with different loss functions. Notably, we see that our novel use of DTW and SIMSE based differentiable loss functions (\DTWEnv{} and \SIMSESpec) can outperform what are regarded as the SOTA loss functions in 3 of 4 cases, although this is highly synthesizer dependent.

P-Loss and MSS have frequently been used as automatic performance measures, yet their ``preference'' has rarely been compared to human rankings. We observed that automatic performance measures and manual listening tests were often in agreement, despite this, manual verification of sound-matching results remains a necessity for future works.

@@

Like nearly all previous work in sound-matching, the main measure of success here was the accurate \textit{replication} of sounds (or synth parameters) rather than imitation of sounds outside of the synthesizer's domain. Replication of arbitrary features of sounds is an important component of sound-design, though it is much harder to define or measure. Future works could explore loss functions which only measure certain characteristics of sound (such as \DTWEnv), and whether they can pave the way for better imitation in sound-matching.

>>>>

\textbf{Q1}: To what extent do automatic evaluation metrics agree with manual listening tests? 
\\We see somewhat consistent results between the rankings assigned by manual hearing tests (which we take as the ground truth), and automatic measures of P-Loss and MSS. With the exception of P-Loss in \BPNoise{} narrowly selecting a different spectrogram based loss, all measures of performance selected the same top performer. However, given the simplicity of our programs and the occasional deviations, we cannot conclude that these automatic measures are substitutes for human hearing tests. As many previous works have done (see Table~\ref{tab:summary}), the use of MSS and/or P-Loss as general loss functions does seem appropriate in general cases, however, the use of listening tests or custom made loss functions in specific contexts remains necessary.

\textbf{Q2}: Are DTW and SIMSE effective measures of loss? If so, in what contexts?\\
\DTWEnv{} consistently outperformed other measures in amplitude-modulated synthesis, where envelope alignment is critical. \SIMSESpec{} performed best in subtractive synthesis, where scale-invariance captures noise-filtering effects more robustly than L1-based measures. These results suggest that the advantages of loss functions are context-dependent, underscoring the need for further exploration of creative differentiable similarity measures.

\textbf{Q3}: Can iterative differentiable optimization be applied effectively with synthesizers using classical DSP functions? \\
The approach of designing DSP functions in Faust and transpiling to differentiable Faust code was convenient, yielded meaningful outputs, and enabled comparative evaluation of losses. We were able to run our 1200 experiments within 72 hours on a laptop without a GPU\footnote{Lenovo ThinkPad T480, i5-8350U CPU at 1.70GHz, 32 GB RAM.}.

\subsection{Practical Recommendations}
    Our findings not only provide answers to the research questions proposed in the introduction, but also serve as guidelines for practitioners in selecting appropriate similarity measures and synthesis methods for future experiments.
\begin{itemize}
    \item In lieu of listening tests, MSS or P-Loss are generally useful for large-scale benchmarking. However, we encourage confirmation of such results with manual listening, particularly if specificity is important. 
    \item For amplitude-modulated synthesis, DTW-based losses such as \DTWEnv{} are recommended over spectrogram differences and JTFS.
    \item  For subtractive/noise-filtered synthesis, spectrogram based losses are most effective. Based on our results, SIMSE appears to be the better method of spectrogram comparisons relative to L1, but the amplitude agnosticism of SIMSE likely plays a role in its poor performance in additive synthesis.  
    \item Iterative differentiable optimization via the Faust-to-JAX pipeline is a viable strategy for defining various differentiable DSP functions, requiring only modest hardware resources and providing a more natural approach to synthesizer definition.
    \item In our amplitude modulation experiments, \JTFS{} did not outperform other measures, contradicting prior claims of its superiority for mesostructures~\cite{vahidi2023mesostructures}. Thus, our results call into question its effectiveness as a general mesostructure measure, and highlight a need for further research into its utility.
\end{itemize}

@@

The main hypothesis tested here is whether the performance of differentiable loss functions is influenced by the synthesis techniques used for sound-matching. We conducted systematic iterative sound-matching experiments by combining four different loss functions with four different sound synthesis programs. We ranked the performance of the iterative sound-matching pipelines for every loss function and program, and observed that the success of the pipeline (that is, how closely the output sound matches the target sound) is program dependent. In other words, different synthesizer programs work best with different loss functions. Notably, we see that our novel use of DTW and SIMSE based differentiable loss functions (\DTWEnv{} and \SIMSESpec) can outperform what are regarded as the SOTA loss functions in 3 of 4 cases, although their success is highly synthesizer dependent.

P-Loss and MSS have frequently been used as automatic performance measures, yet their ``preference'' has rarely been compared to human rankings. We observed that automatic performance measures and manual listening tests were generally in agreement, despite this, manual verification of sound-matching results remains a necessity in non-general experiments due to occasional divergences between automatic and manual tests.

@@

Like nearly all previous work in sound-matching, the main measure of success here was the accurate \textit{replication} of sounds (or synthesizer parameters) rather than \textit{imitation} of sounds outside of the synthesizer's domain. Replication of arbitrary features of sounds is an important component of sound-design, though it is much harder to define or measure. Future works could explore loss functions which only measure certain characteristics of sound (such as \DTWEnv), and whether they can pave the way for better imitation in sound-matching.
